Metric Spaces with Expensive Distances

### 2018-12-05 paper/draft.tex
Spelling

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
an $\eps$-spanner immediately. Net-trees can also be used to compute approximate nearest neighbors
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylog time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances that have been computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see Figure~\ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
operations whenever a new distance has been computed, turing the method useless in the standard computational model.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most succesful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
\michael{Acutally, is that true? We could run BlindGreedy on the 2-approximations instead of on $[0,\infty]$,
that might save something}
We also compare with a spanner construction based on WSPDs. Our simple algorithms tend to give much smaller
spanners on the tested example. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

\michael{Add references here whereever appropriate}
The greedy spanner is guaranteed (\cite{althofer1993sparse}) to return a spanner of size $O(n)$
(for constant doubling dimension and fixed $\eps>0$); in experimental study \cite{farshi2009experimental}
it was also shown to return the sparsest graph. However, it clearly
has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
this means that in our cost model, the greedy spanner has the worst possible
performance.

On the other hand, spanner constructions based on WSPDs only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct an $(1+\eps)$-spanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

@@

\begin{description}
\item[\textsc{BlindRandom}] Among all pairs $(i,j)$ where $\frac{b_{i,j}}{a_{i,j}}>(1+\eps)$,
we pick one pair uniformly at random
\item[\textsc{BlindGreedy}] Pick the pair $(i,j)$ which maximizes the ratio $\frac{b_{i,j}}{a_{i,j}}$.
If the maximizing pair is not unique, choose among the maximizing pairs uniformly at random.
\end{description}
%
The idea behind \textsc{BlindGreedy} is that we query an edge for which we know the least,
in that way hoping to gather most additional information about the metric space.
Also, our conventions imply that in \textsc{BlindGreedy} the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
lower bounds for every edge before it will start adding any other edges. Based on this observation,
we also tested variations of the \textsc{BlindRandom} algorithm, where the algorithm
first enforces connectedness and/or lower bounds (i.e., if there are infinite upper bounds,
then the algorith can only choose one of the corresponding edges, etc).

@@

\textsc{BlindQuasiSortedGreedy} tries to mimick the blind greedy spanner and hence appears as a natural
choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
results. The reason is that no pair acquires useful lower bounds when only short distance are queried
(the greedy spanner does not have this issue because it knows the distance and hence does not need
lower bounds).
Generally speaking, short distances are good for sharp upper bounds, whereas long distances are
useful for lower bounds. This motivates \textsc{BlindQuasiSortedShaker}
which alternates between short and long distances.

@@

We point out that the proof fails for $\eps=0$ because in that case, we cannot exlude an $\eps$-ball
of close-by points as in the second part of Lemma~\ref{lem:bound_lemma}, and the packing argument fails.
Indeed, as the example in Figure~\ref{fig:exact_bad_example} shows, there are point sets where the expected number
of distance computations for exact nearest neighbor is linear.

@@

%\paragraph{Notion of Dimensionality.}
%%
%Exansion constant, doubling dimension.

@@

%Suppose that the pair $(a_1, b_1)$ with $a_1 \in A$, $b_1 \in B$ was processed and the edge $(a, b)$
%was added to the spanner. Then from claim 1 we know that we already have a $t$-spanner on $A$ and $B$.
%Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been addedhttps://v2.overleaf.com/project/5bf6fa26032afa24d94383a5
%to the spanner. By claim 1
%$d_G(a_2, a_1) \leq td(a_2, a_1)$ and $d_G(b_1, b_2) \leq t d(b_1, b_2)$.
%By properties of WSPD, $d_G(a_1, b_1) = d(a_1, b_1) \leq (1 + \frac{2}{s})d(a_2, b_2)$
%and $t d(a_2, a_1) \leq \frac{t}{s} d(a_2, b_2)$, $t d(b_1, b_2) \leq \frac{t}{s} d(a_2, b_2)$.

@@

\section{Conclusion and future work}
\label{sec:conclusion}
%
We have introduced a new cost model for the analysis of algorithms
for metric spaces that fits the situation that computing an individual distance
is more costly than other types of primitive operations.
Our theoretical and experimental results are under the usual assumption
that the metric space has a low doubling dimension.
However, in our motivating example of collections of persistence diagrams
or Reeb graphs, this assumption does not hold. For instance,
the space of persistence diagrams has an infinite doubling dimension.
Nevertheless, realisitic data sets are usually not just a random sample
in that infinite-dimensional space, but have structures
(e.g. clusters of close-by diagrams) which should be favorable for our approach
We plan to consider the quality of our algorithms for persistence diagrams
as future work.

On the theoretical side, the obvious next question is whether our strategy
for blind spanners yields a linear spanner in expectation. 
Our experiments are not conclusive enough in this respect to make this conjecture yet.
However, it has been brought to our attention\footnote{Yusu Wang, personal communication}
that the size of the blind spanner is bounded by the \emph{weight} of the WSPD
which is the sum of the cardinalities of all pairs in a WSPD.
The weight of a WSPD can be quadratic, but premilinary
experimental evaluation on worst-case examples do not show such a quadratic
behavior. Therefore, we postpone the theortical analysis of the spanner construction
to an extended version of this article.

@@

\ignore{
We presented experimental evidence showing that in low
doubling dimension we can avoid many distance computations,
if we try to get as much information as possible from the triangle inequality
applied to the previously computed distance. The \textsc{BlindGreedy} spanner
performs especially well, being able to produce  sparse spanners even for small values of  $\eps$.
For the approximate nearest neighbor algorithm we proved a logarithmic upper bound,
and, even though the constant in the proof was very large, we demonstrated
experimentally that this does not show in practice.

The next step on the theoretical side could be proving some upper bound
for blind spanners; it is also interesting to look at other metric space
algorithms and data structures in the setting of expensive distance computation.
On the more applied side, the further development of this approach would
be to find finite spaces of low dimension in practically
important cases, like spaces of persistence diagrams or images endowed with Wasserstein (a.k.a. Earth Movers)
metric. We know that low-dimensional subspaces of functions do not necessarily produce
low-dimensional spaces of persistence diagrams, but there may be some conditions
which guarantee that dimension does not increase too much.
% An easy example: consider a fixed function $f$ to which we apply
% a finite-dimensional family of perturbations, $f(x) + g(x; \lambda_1, \dots, \lambda_n)$;
% for instance, each $\lambda_i$ is associated with some $x_i$ and adds a Gaussian noise
% scaled by $\lambda$.
% If the perturbations are very small, then they will produce some small wiggles on the function,
% and then the optimal matcih
}

>>>>

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
an $\eps$-spanner immediately. Net-trees can also be used to compute approximate nearest neighbors
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

@@

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances that have been computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see Figure~\ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
operations whenever a new distance has been computed, turning the method useless in the standard computational model.

@@

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested example. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

The greedy spanner is guaranteed (\cite{althofer1993sparse}) to return a spanner of size $O(n)$
(for constant doubling dimension and fixed $\eps>0$); in experimental study \cite{farshi2009experimental}
it was also shown to return the sparsest graph. However, it clearly
has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
this means that in our cost model, the greedy spanner has the worst possible
performance.

On the other hand, spanner constructions based on WSPD only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct an $(1+\eps)$-spanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

@@

\begin{description}
\item[\textsc{BlindRandom}] Among all pairs $(i,j)$ where $\frac{b_{i,j}}{a_{i,j}}>(1+\eps)$,
we pick one pair uniformly at random
\item[\textsc{BlindGreedy}] Pick the pair $(i,j)$ which maximizes the ratio $\frac{b_{i,j}}{a_{i,j}}$.
If the maximizing pair is not unique, choose among the maximizing pairs uniformly at random.
\end{description}
%
The idea behind \textsc{BlindGreedy} is that we query an edge for which we know the least,
in that way hoping to gather most additional information about the metric space.
Also, our conventions imply that in \textsc{BlindGreedy} the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
lower bounds for every edge before it will start adding any other edges. Based on this observation,
we also tested variations of the \textsc{BlindRandom} algorithm, where the algorithm
first enforces connectedness and/or lower bounds (i.e., if there are infinite upper bounds,
then the algorithm can only choose one of the corresponding edges, etc).

@@

\textsc{BlindQuasiSortedGreedy} tries to mimic the blind greedy spanner and hence appears as a natural
choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
results. The reason is that no pair acquires useful lower bounds when only short distance are queried
(the greedy spanner does not have this issue because it knows the distance and hence does not need
lower bounds).
Generally speaking, short distances are good for sharp upper bounds, whereas long distances are
useful for lower bounds. This motivates \textsc{BlindQuasiSortedShaker}
which alternates between short and long distances.

@@

We point out that the proof fails for $\eps=0$ because in that case, we cannot exclude an $\eps$-ball
of close-by points as in the second part of Lemma~\ref{lem:bound_lemma}, and the packing argument fails.
Indeed, as the example in Figure~\ref{fig:exact_bad_example} shows, there are point sets where the expected number
of distance computations for exact nearest neighbor is linear.

@@

%\paragraph{Notion of Dimensionality.}
%%
%Expansion constant, doubling dimension.

@@

%Suppose that the pair $(a_1, b_1)$ with $a_1 \in A$, $b_1 \in B$ was processed and the edge $(a, b)$
%was added to the spanner. Then from claim 1 we know that we already have a $t$-spanner on $A$ and $B$.
%Consider another pair $(a_2, b_2)$. Let $G$ be the weighted graph we have immediately after the edge $(a_1, b_1)$ has been added
%to the spanner. By claim 1
%$d_G(a_2, a_1) \leq td(a_2, a_1)$ and $d_G(b_1, b_2) \leq t d(b_1, b_2)$.
%By properties of WSPD, $d_G(a_1, b_1) = d(a_1, b_1) \leq (1 + \frac{2}{s})d(a_2, b_2)$
%and $t d(a_2, a_1) \leq \frac{t}{s} d(a_2, b_2)$, $t d(b_1, b_2) \leq \frac{t}{s} d(a_2, b_2)$.

@@

\section{Conclusion and future work}
\label{sec:conclusion}
%
We have introduced a new cost model for the analysis of algorithms
for metric spaces that fits the situation that computing an individual distance
is more costly than other types of primitive operations.
Our theoretical and experimental results are under the usual assumption
that the metric space has a low doubling dimension.
However, in our motivating example of collections of persistence diagrams
or Reeb graphs, this assumption does not hold. For instance,
the space of persistence diagrams has an infinite doubling dimension.
Nevertheless, realistic data sets are usually not just a random sample
in that infinite-dimensional space, but have structures
(e.g. clusters of close-by diagrams) which should be favorable for our approach
We plan to consider the quality of our algorithms for persistence diagrams
as future work.

On the theoretical side, the obvious next question is whether our strategy
for blind spanners yields a linear spanner in expectation. 
Our experiments are not conclusive enough in this respect to make this conjecture yet.
However, it has been brought to our attention\footnote{Yusu Wang, personal communication}
that the size of the blind spanner is bounded by the \emph{weight} of the WSPD
which is the sum of the  cardinalities of all pairs in a WSPD.
The weight of a WSPD can be quadratic, but preliminary
experimental evaluation on worst-case examples do not show such a quadratic
behavior. Therefore, we postpone the theoretical analysis of the spanner construction
to an extended version of this article.

@@
